\begin{abstract}
    Population models are mathematical models to study the dynamics of a population. Markov
    population processes are Markov chains of continuous or discrete time, in which a state tracks
    population size of each of the involved species. We study a framework for data-informed
    parameter synthesis of parametric Markov population processes. Given experimental data for the
    population at its steady-state, parameter synthesis aims to find a set of parameters satisfying
    a temporal property of interest. We assume that experimental observations of the population are
    available at steady-state. We design and compare performance of a Bayesian framework for
    parameter synthesis in two cases: (i) when the exact likelihood function of the property of
    interest is computed in a pre-processing step, and (ii) when it has to be approximated by means
    of Monte-Carlo methods (it is likelihood-free). The frameworks are constructed with different
    sampling and optimization techniques to approximate the posterior distribution. We evaluate the
    frameworks using various population models of different sizes, using synthetic data generated
    from a known true parameter. By measuring the distance between estimated parameters and true
    parameters, and by visualize synthesized parameter values with their corresponding posterior
    estimate, we show that both frameworks can derive a set of satisfying parameter values and an
    estimation that is close to the true parameter. Furthermore, results show that the
    likelihood-free framework is overall performing better.
\end{abstract}

