\chapter{Related works}
%% Frameworks from molyneux
The frameworks presented in this thesis are based on ABC-SMC framework \cite{molyneux2019bayesian}
and ABC-(SMC)2 \cite{molyneux2020abc} by Molyneux et al. However, the ABC-SMC and ABC-(SMC)2
frameworks synthesize parameters for CTMC and check the CTMC model agains CSL property. In
parametric DTMC, since the symbolic rational function of PCTL property is obtainable, we based on
Del Moral \cite{del2006sequential} and Daviet \cite{daviet2018inference} to construct an algorithm
based on evaluation of symbolic rational function, then benchmark it againts the approach based on
only simulation.

%% Model checking
The theoretical background of model checking DTMC is presented by Baier et al
\cite{baier2008principles}. An introduction to parametric DTMC and current methods on parameter
synthesis is presented by the tutorial by Katoen \cite{katoen2016probabilistic}. More in-depth
surveys and discoveries on parametric model checking and parameter synthesis is presented by Junges
\cite{junges2020parameter} and \cite{hutschenreiter2017parametric}. Polgreen
\cite{polgreen2016data}, and Haesaert \cite{haesaert2015data} present Bayesian approach on
data-informed parameter synthesis and model verification. Jha \cite{jha2009bayesian} pioneered the
first algorithm for Bayesian approach on Statistical Model Checking, namelyBayesian Sequential
Hypothesis Testing.

%% Bayesian inference: optimzation methodds
Monte-Carlo sampling algorithms used in this thesis are presented by Metropolis
\cite{metropolis1953equation}, Del Moral \cite{del2006sequential}, and Toni
\cite{toni2009approximate}. A comparision between different Monte-Carlo sampling algorithms,
including Markov-chain Monte-Carlo, Sequential Monte-Carlo, Hamiltonian Monte-Carlo, and Hamiltonian
Sequential Monte-Carlo is presented in \cite{daviet2018inference}. Silk \cite{silk2012optimizing}
discussed different approaches on the pertubation kernel selection of Sequential Monte-Carlo and
Sequential Monte-Carlo with Approximate Bayesian Computation algorithms.

%% Tools: PRISM and STORM
The model checking step in the frameworks presented by this thesis are implemented using Storm model
checker \cite{hensel2020probabilistic}. Storm provides well documented and easy to use APIs to embed
model checking to software projects programmatically. However, Storm does not support Statistical
Model Checking. Thus, Statistical Model Checking step in simulation-based frameworks is implemented
using PRISM \cite{kwiatkowska2011prism}.